---
title: "Data Transformation with `dplyr` and `pandas`"
author: "Kevin Liu"
date: "02/29/2020"
header-includes:
  - \usepackage{upgreek}
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 4
    self_contained: true
bibliography: "bibliography.bib"
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.align = "center")
```

I use two popular package `dplyr` in `r` and `pandas` in `python` to accomplish same data transformation goals. Hope this webpage could help anyone who are learning `R` from `Python` or `Python` from `R`.

# load nycflights 13
To demonstrate data management in both packages, I use `nycflights13::flights`.

R code
```{r nycflights_r}
library(tidyverse)
library(knitr)
library(nycflights13)
data(flights)
head(as_tibble(flights))
##only run following code for once for sake of time
##write.csv(flights,"flights.csv",row.names = FALSE)
```

Python Code
```{python nycflights_y}
import numpy as np
import pandas as pd
df1 = pd.read_csv("flights.csv")
df1.head()
```

# Filter rows

## AND operator
R Code
```{r and_r}
flights %>% filter(month == 1, day == 1)
```

Python Code
```{python and_p}
filter_row = (df1['month'] == 1) & (df1['day'] == 1)
df1[filter_row].head()
```

## IN operator

R code
```{r in_r}
flights %>% filter(month %in% c(11,12))
```

Python Code
```{python in_p}
filter_row = df1['month'].isin([11,12])
df1[filter_row].head()
```

# Sort Rows
R code
```{r sort_r}
flights %>% arrange(desc(arr_delay))
```

Python Code
```{python sort_p}
df1.sort_values(by=['arr_delay'],ascending=False).head()
```

# Select column

R code

```{r select_col_r}
flights %>% select(year:day)
```
Python Code

```{python select_col_y}
df1.loc[:,"year":"day"].head()
```

# unselect column

R code

```{r unselect_r}
flights %>% select(-(year:day))
```

Python Code

```{python unselect_y}
sel_col = df1.loc[:,"year":"day"].columns
sel_col = ~np.isin(df1.columns,sel_col)
df1.loc[:,sel_col].head()
```

`r if (knitr::is_html_output()) '# References {-}'`
@R_for_data_science