---
title: "Example for Mixed Model Equations and REML"
author: "Kevin Liu"
date: "7/10/2018"
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 4
    self_contained: true
bibliography: "bibliography.bib"
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, fig.align='center')
```

I will use `bond` data from Chapter2 of @Littell:2006:SMM:1205543 as a example.

#Load Packages
```{r}
library(lme4)
library(optimization)
```

# Create Dataframe `df.bond`

```{r}
df.bond <- data.frame(ingot = rep(1:7,each = 3),
                      metal = rep(c("n","i","c"),7),
                      pres = c(67,71.9,72.2,
                               67.5,68.8,66.4,
                               76,82.6,74.5,
                               72.7,78.1,67.3,
                               73.1,74.2,73.2,
                               65.8,70.8,68.7,
                               75.6,84.9,69))
df.bond$ingot <- factor(df.bond$ingot)
df.bond
```

# Linear mixed model by maximum likelihood estimation

First we build linear mixed model by `lme4` package with MLE method. Then we compare the parameter estimations by using equations from [Some important proof of Mixed Model Equations and REML](Linear_mixed_model_equations.html) via `optimization` package.

```{r}
lmm.bond <- lmer( pres ~ metal + (1|ingot),REML = FALSE,data = df.bond)
summary(lmm.bond)
anova(lmm.bond)
```

For linear mixed model

\begin{equation}
\boldsymbol{Y} = \boldsymbol{X\beta} + \boldsymbol{Zu} + \boldsymbol{e}
\end{equation}

\begin{equation}
\begin{pmatrix}
\boldsymbol{u}\\
\boldsymbol{e}
\end{pmatrix} \sim \mathcal{N}
\begin{pmatrix}
\begin{pmatrix}
\boldsymbol{0}\\
\boldsymbol{0}
\end{pmatrix},
\begin{pmatrix}
\boldsymbol{G} & \boldsymbol{0}\\
\boldsymbol{0} & \boldsymbol{R}
\end{pmatrix}
\end{pmatrix}
\end{equation}

Design matrix $\boldsymbol{X}$ is

```{r}
matrix.x <- model.matrix(lmm.bond)
matrix.x
```

Design matrix $\boldsymbol{Z}$ is 

```{r}
df.z <- read.csv("files/Z.csv", header = FALSE)
matrix.z <- as.matrix(df.z)
colnames(matrix.z) <- paste0("ingot:",1:7)
matrix.z
```

Define matrix $\boldsymbol{G}$ as

$$\boldsymbol{G}_{7 \times 7} = \boldsymbol{I} \sigma^2_{\text{ingot}}$$

Define matrix $\boldsymbol{R}$ as

$$\boldsymbol{R}_{21 \times 21} = \boldsymbol{I} \sigma^2_{\text{residual}}$$

Use equation(4.9) and equation (6.1) from [Some important proof of Mixed Model Equations and REML](Linear_mixed_model_equations.html)

\begin{equation}
\tilde{\boldsymbol{\beta}} 
= 
(\boldsymbol{X}^T \boldsymbol{V}^{-1} \boldsymbol{X})^{-1} 
\boldsymbol{X}^T \boldsymbol{V}^{-1} \boldsymbol{Y}
\tag{4.9}
\end{equation}

\begin{equation}
-2\ell(\boldsymbol{\theta};\boldsymbol{Y})
=
\text{log}|\tilde{\boldsymbol{V}}(\boldsymbol{\theta})| 
+ 
(\boldsymbol{Y}-\boldsymbol{X}\tilde{\boldsymbol{\beta}}(\boldsymbol{\theta}))^T
\tilde{\boldsymbol{V}}(\boldsymbol{\theta})^{-1}
(\boldsymbol{Y}-\boldsymbol{X}\tilde{\boldsymbol{\beta}}(\boldsymbol{\theta}))
\tag{6.1}
\end{equation}


Estimation of covariance parameters
```{r}
loglikef <- function(x) {
  vector.Y <- as.vector(df.bond$pres)
  matrix.G <- x[1] * diag(1,nrow = 7)
  matrix.R <- x[2] * diag(1,nrow = 21)
  matrix.V <- matrix.z %*% matrix.G %*% t(matrix.z) + matrix.R
  vector.Beta <- solve(t(matrix.x) %*% solve(matrix.V) %*% matrix.x) %*% t(matrix.x) %*% solve(matrix.V) %*% vector.Y
  loglike <- log(det(matrix.V)) + t(vector.Y - matrix.x %*% vector.Beta) %*% solve(matrix.V) %*% (vector.Y - matrix.x %*% vector.Beta)
  return(c(loglike))
}

optim_nm(fun = loglikef, k = 2, start = c(1,1),maximum= FALSE, tol = 0.0000000000001)$par
hat.cov.para <- optim_nm(fun = loglikef, k = 2, start = c(1,1),maximum= FALSE, tol = 0.0000000000001)$par
```

We have that 

$$\sigma^2_{\text{ingot}} = 9.812383 \text{ and } \sigma^2_{\text{residual}} = 8.889932$$
This is identical to estimations from `lme4`.

Then estimate fix effect $\boldsymbol{\beta}$

```{r}
hat.beta <- function(x) {
  vector.Y <- as.vector(df.bond$pres)
  matrix.G <- x[1] * diag(1,nrow = 7)
  matrix.R <- x[2] * diag(1,nrow = 21)
  matrix.V <- matrix.z %*% matrix.G %*% t(matrix.z) + matrix.R
  vector.Beta <- solve(t(matrix.x) %*% solve(matrix.V) %*% matrix.x) %*% t(matrix.x) %*% solve(matrix.V) %*% vector.Y
  return(vector.Beta)
}
round(hat.beta(x = hat.cov.para),4)
```

We have that 
$$\boldsymbol{\beta}
=
\begin{bmatrix} \beta_0\\
\beta_1\\
\beta_2
\end{bmatrix} =
\begin{bmatrix}
70.1857\\
5.7143\\
0.9143
\end{bmatrix}$$

Again, this is identical to estimations from `lme4`.

# Linear mixed model by REML

Same as [section 3][Linear mixed model by maximum likelihood estimation], firstly we build linear mixed model by `lme4` package with REML option. Then we compare 'hand' calculated estimations with estimations from `lme4`.

```{r}
relmm.bond <- lmer( pres ~ metal + (1|ingot),REML = TRUE,data = df.bond)
summary(relmm.bond)
```

Using equation(7.3) from [Some important proof of Mixed Model Equations and REML](Linear_mixed_model_equations.html)

\begin{equation}
-2\ell_{REML}(\boldsymbol{\theta};\boldsymbol{Y}) = 
\text{log} |\tilde{\boldsymbol{V}}(\boldsymbol{\theta})|
+
\text{log} |\boldsymbol{X}^T \tilde{\boldsymbol{V}}^{-1} \boldsymbol{X}| 
+
(\boldsymbol{Y} - \boldsymbol{X}\tilde{\boldsymbol{\beta}}(\boldsymbol{\theta}))^T
\tilde{\boldsymbol{V}}^{-1}
(\boldsymbol{Y} - \boldsymbol{X}\tilde{\boldsymbol{\beta}}(\boldsymbol{\theta}))
\tag{7.3}
\end{equation}

```{r}
reloglikef <- function(x) {
  vector.Y <- as.vector(df.bond$pres)
  matrix.G <- x[1] * diag(1,nrow = 7)
  matrix.R <- x[2] * diag(1,nrow = 21)
  matrix.V <- matrix.z %*% matrix.G %*% t(matrix.z) + matrix.R
  vector.Beta <- solve(t(matrix.x) %*% solve(matrix.V) %*% matrix.x) %*% t(matrix.x) %*% solve(matrix.V) %*% vector.Y
  loglike <- log(det(matrix.V)) + log(det(t(matrix.x) %*% solve(matrix.V) %*% matrix.x)) + t(vector.Y - matrix.x %*% vector.Beta) %*% solve(matrix.V) %*% (vector.Y - matrix.x %*% vector.Beta)
  return(c(loglike))
}

optim_nm(fun = reloglikef, k = 2, start = c(5,6),maximum= FALSE, tol = 0.0000000000001)$par
```

Again, parameter estimations are close enough. 

`r if (knitr::is_html_output()) '# References {-}'`
@Littell:2006:SMM:1205543